---
title: "R2camtrapdp: schema-driven workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R2camtrapdp: schema-driven workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(R2camtrapdp)
```

# Overview

`R2camtrapdp` converts camera-trap data held in an arbitrary spreadsheet into a
[Camera Trap Data Package (Camtrap DP)](https://camtrap-dp.tdwg.org/).

This version is **schema-driven**: the structure, types and constraints of the
output tables are read from the official Frictionless *table schemas* of the
Camtrap DP version you choose. As a result the package

* works with **any Camtrap DP version** (`1.0`, `1.0.1`, `1.0.2`) — and with
  other **schema flavors** such as the bioacoustics extension (see §8) — simply
  by pointing it at the right schema,
* understands **custom / extra columns** that a particular schema (or your own
  project) defines,
* checks every value against the schema constraints
  (`required`, `unique`, `enum`, `minimum`/`maximum`, `pattern`, date/datetime
  **format**, ...),
* checks the **relations** between tables (primary keys and foreign keys),
* surfaces every **URL-specified reference** in a schema (semantic mappings and
  description links) so nothing is overlooked (see §1), and
* can run the Python **Frictionless** validator and report the errors back in R.

The classic helper functions (`create_deployments()`, `create_media()`,
`create_observations()`) and the `R6_CamtrapDP` class keep the same names and
arguments as before, so existing scripts continue to work. The new
schema-driven behaviour is added on top.

> **Note on internet access.** Setting a table (`set_deployments()` etc.)
> downloads the table schema for the chosen `version` from GitHub the first time
> it is needed, and then caches it. If you work offline, pass a downloaded
> schema file with the `local_schema =` argument.

# Data

The package ships with example data for several deployments with image records.

```{r}
# multiple deployments with image data
data("Idep")   # deployment table
data("Iobs")   # observation table
```

`Idep` holds one row per deployment (camera placement) with columns such as
`deploymentID`, `longitude`, `latitude`, `locationID`, `startDate`/`startTime`,
`endDate`/`endTime`, `cameraID`, `cameraModel`, `Delay`, `Height`, `bait` and
`setupBy`. `Iobs` holds one row per observation with the institution/collection
codes, `filename`, `deploymentID`, `date`/`time`, `obsID`, `eventID`,
`eventStart`/`eventEnd`, `object`, `genus`, `species`, `class` and
`individualCount`.

# 1. Choose a version and inspect its schema (optional)

The whole pipeline is driven by the schema of the version you pick. Camtrap DP
versions `1.0`, `1.0.1` and `1.0.2` are all supported; their table schemas share
the same field names, types and constraints, so the only practical difference is
that `1.0.2` recognises a few more missing-value tokens (`NA`, `NaN`, `nan`). You
can inspect the schema of any version directly with `TableSchema`.

(Note: the official `1.0` *profile* — the metadata JSON Schema — has an upstream
bug, a malformed internal `$ref`, that newer Frictionless rejects. Specifying
`version = "1.0"` therefore emits a warning; `validate_frictionless()` works
around the bug automatically, but `1.0.1` or later is recommended.)

```{r, eval = FALSE}
version <- "1.0.1"

dep_schema <- TableSchema$new("deployments", version = version)
dep_schema$field_names()           # every column the schema defines
dep_schema$required_field_names()  # columns that must be present and non-missing
dep_schema$empty_table()           # a 0-row, correctly typed "shell" table
```

You rarely need to do this by hand — the `R6_CamtrapDP` object loads and caches
the right schema for you — but it is useful for understanding what a given
version expects.

`check_schema()` confirms that the schema itself is a well-formed Frictionless
Table Schema (supported field `type`s, constraints that are valid for each type,
primary/foreign keys that reference defined fields) — useful before adopting a
brand-new or hand-edited schema.

```{r, eval = FALSE}
dep_schema$check_schema()
```

## External (URL) references in a schema

Some Camtrap DP information is specified not as a machine-checkable constraint
but as a **URL**: semantic mappings (`skos:exactMatch` / `broadMatch` /
`narrowMatch` to Darwin Core, Audubon Core, ... terms) and reference URLs in
field descriptions (for example the IANA media-type registry for
`fileMediatype`, or method DOIs for `individualSpeed`). The package only enforces
the structured constraints; the URL-referenced meaning is *not* validated. To
make sure you never overlook such a specification when adopting a version or a
new schema flavor, list them with:

```{r, eval = FALSE}
dep_schema$external_references()   # every URL the schema declares (skos, descriptions, schema URL)
dep_schema$semantic_only_fields()  # fields whose meaning is URL-defined and cannot be value-checked
```

`external_references()` returns a tidy table (`resource`, `field`, `key`,
`category`, `url`); `semantic_only_fields()` flags the columns you should check
against the referenced authority by hand. The whole package can be scanned at
once with `datapackage$external_references()`.

# 2. Build the three core tables

## Create deployments

Using the deployment data (`Idep`), the deployments table is created exactly as
before. `create_deployments()` accepts either combined datetimes or separate
date/time columns.

```{r}
deployments <- create_deployments(
  deploymentID         = Idep$deploymentID,
  longitude            = Idep$longitude,
  latitude             = Idep$latitude,
  locationID           = Idep$locationID,
  deploymentStart_date = Idep$startDate,
  deploymentStart_time = Idep$startTime,
  deploymentEnd_date   = Idep$endDate,
  deploymentEnd_time   = Idep$endTime,
  cameraID             = Idep$cameraID,
  cameraModel          = Idep$cameraModel,
  cameraDelay          = Idep$Delay,
  cameraHeight         = Idep$Height,
  baitUse              = Idep$bait,
  setupBy              = Idep$setupBy)
```

`create_deployments()` also accepts (not shown above): `deploymentStart` /
`deploymentEnd` (combined datetimes, used instead of the `*_date` / `*_time`
pairs), `locationName`, `coordinateUncertainty`, `cameraDepth` (mutually
exclusive with `cameraHeight`), `cameraTilt`, `cameraHeading`,
`detectionDistance`, `timestampIssues`, `featureType`, `habitat`,
`deploymentGroups`, `deploymentTags`, `deploymentComments`, and `tz` (time zone,
default `"Japan"`).

## Create media

```{r}
# media ID
mediaIDi <- paste(Iobs$institutionCode,
                  Iobs$collectionCode,
                  Iobs$locationID,
                  as.numeric(factor(Iobs$filename)),
                  sep = "_")

# file information
fileName      <- Iobs$filename
filetype      <- tolower(unlist(lapply(strsplit(fileName, "\\."), "[", 2)))
fileMediatype <- paste("image", filetype, sep = "/")
filePublic    <- !grepl("ヒト", fileName)   # hide human images from the public

media <- create_media(
  mediaID        = mediaIDi,
  deploymentID   = Iobs$deploymentID,
  timestamp_date = Iobs$date,
  timestamp_time = Iobs$time,
  filePath       = "Image",
  filePublic     = filePublic,
  fileMediatype  = fileMediatype,
  captureMethod  = "activityDetection",
  fileName       = fileName)
```

`create_media()` also accepts (not shown above): `timestamp` (combined datetime,
instead of `timestamp_date` / `timestamp_time`), `exifData`, `favorite`,
`mediaComments`, `tz`, and `omitduplicate` (drop duplicate `mediaID`s, default
`TRUE`).

## Create observations

```{r}
# event-based observations
observationLevel <- "event"

# observationType must be one of the schema enum values
observationType <- ifelse(Iobs$object == "hito", "human",
                   ifelse(Iobs$object == "none", "blank",
                   ifelse(Iobs$object == "unidentifiable", "unknown", "animal")))

# scientific name
scientificName <- ifelse(is.na(Iobs$genus), Iobs$class, paste(Iobs$genus, Iobs$species))

# unique observation IDs
observationID <- paste(mediaIDi, Iobs$obsID, sep = "_")

observations <- create_observations(
  observationID             = observationID,
  deploymentID              = Iobs$deploymentID,
  eventID                   = Iobs$eventID,
  eventStart                = Iobs$eventStart,
  eventEnd                  = Iobs$eventEnd,
  observationLevel          = observationLevel,
  observationType           = observationType,
  scientificName            = scientificName,
  count                     = Iobs$individualCount,
  classificationMethod      = "human",
  classificationProbability = 1)
```

`create_observations()` also accepts (not shown above): `mediaID`, the
`eventStart_date` / `eventStart_time` and `eventEnd_date` / `eventEnd_time` pairs
(instead of combined `eventStart` / `eventEnd`), `cameraSetupType`, `lifeStage`,
`sex`, `behavior`, `individualID`, `individualPositionRadius`,
`individualPositionAngle`, `individualSpeed`, `bboxX`, `bboxY`, `bboxWidth`,
`bboxHeight`, `classifiedBy`, `classificationTimestamp`, `observationTags`,
`observationComments`, `tz`, and `omitduplicate`.

# 3. Assemble the data package

## Create the R6 object (with a version)

```{r}
datapackage <- R6_CamtrapDP$new(version = "1.0.1")
```

The `version` you give here selects the schemas used for validation and written
into `datapackage.json`. Change it to target a different Camtrap DP release.

## Import the tables (now schema-validated)

`set_deployments()`, `set_media()` and `set_observations()` keep their original
names, but now each one **coerces the table to the schema types and validates it
against the schema** for the chosen version. Any problems are printed as a
summary; you can switch the printing off with `validate = FALSE`.

```{r, eval = FALSE}
datapackage$set_deployments(deployments)
datapackage$set_media(media)
datapackage$set_observations(observations)
```

*(The chunks that download a schema, write files, look up taxonomy, or call
Python are shown but not executed when this vignette is built, so they produce
no output here.)*

The validation summary tells you, for every issue, the file, the column, the
row, the violated rule and a message — for example a value that breaks an
`enum`, a number outside its `minimum`/`maximum`, or a datetime that does not
match the required format. A value that does not even fit the column type (e.g.
a non-numeric string in a `number` field) is reported as a `type` error rather
than being silently turned into `NA`.

## Check relations between tables

Foreign keys (e.g. `media.deploymentID` must exist in `deployments`, and
`observations.mediaID` must exist in `media`) and primary-key uniqueness are
read from each table's schema and checked across the tables you have added.

```{r, eval = FALSE}
datapackage$check_relations()
```

If a primary-key or a required foreign-key column is **entirely missing** in a
stored table (often a column-name mismatch that coercion filled with `NA`),
`check_relations()` warns and points at the data, e.g.
`datapackage$data$observations has 'deploymentID' entirely missing ...`, so you
can inspect `datapackage$data$<resource>` directly.

# 4. Metadata

Camtrap DP requires five metadata properties (contributors, project, spatial,
temporal, taxonomic — plus `created`). Six further properties are optional. The
metadata functions are unchanged from previous versions.

## Check which metadata the profile requires

The required metadata is itself read from the package **profile** (a JSON
Schema). `metadata_requirements()` lists every required top-level property, the
method that sets it, and whether it is currently set; `check_metadata()`
validates the current object against the profile and reports anything missing
(including nested keys such as `project.samplingDesign`).

```{r, eval = FALSE}
datapackage$metadata_requirements()   # checklist: property, required, set_with, currently_set
datapackage$check_metadata()          # report missing required metadata
```

This is the R-side counterpart of the metadata (profile) validation that
Frictionless performs (§6), so you can confirm the required structure *before*
writing the package and calling Python.

## Required metadata

### Contributors

`add_contributors()` imports a data frame with columns `title`, `email`, `path`,
`role` and `organization`. `role` may be `contact`, `principalInvestigator`,
`rightsHolder`, `publisher` or `contributor`.

```{r}
cd <- data.frame(
  title        = c("Keita Fukasawa", "Kana Terayama"),
  email        = c("fukasawa@nies.go.jp", "terayama.kana@nies.go.jp"),
  path         = c("https://orcid.org/0000-0003-0272-9180",
                   "https://orcid.org/0000-0001-6935-7233"),
  role         = c("contact", "principalInvestigator"),
  organization = c("National Institute for Environmental Studies (NIES)",
                   "National Institute for Environmental Studies (NIES)"))
datapackage$add_contributors(cd)
```

### Project

```{r}
datapackage$set_project(
  title            = "DummyData",
  samplingDesign   = "simpleRandom",
  captureMethod    = "activityDetection",
  individualAnimals = FALSE,
  observationLevel = "event")
```

`samplingDesign` is one of `simpleRandom`, `systematicRandom`,
`clusteredRandom`, `experimental`, `targeted` or `opportunistic`;
`captureMethod` is `activityDetection` or `timeLapse`; `observationLevel` is
`media` or `event`. The optional `id`, `acronym`, `description` and `path`
arguments are also available.

### Spatial and temporal

`set_st()` derives the spatial and temporal coverage from the deployments, so it
must be called after `set_deployments()`.

```{r, eval = FALSE}
datapackage$set_st()
```

### Taxonomic

`set_taxon()` lists the unique `scientificName` values from the observations and
looks up `taxonID`, `taxonRank` and the higher taxonomy from a taxonomic
database (`gbif` by default; also `itis` / `ncbi`; see `taxadb::get_ids`). The
Camtrap DP `taxonomic` block requires a `taxonID` (a GBIF / IUCN identifier or
URI), so `taxadb` is a required dependency of R2camtrapdp (installed with it);
this step also needs internet access.

```{r, eval = FALSE}
datapackage$set_taxon()
```

Names that cannot be matched get `taxonID = NA` (omitted from the output, not a
bogus `<uri>NA`). `set_taxon()` warns about `scientificName` values with
unnecessary whitespace and about names with no `taxonID` in the chosen database,
so you can clean or check those names.

### Created

```{r}
datapackage$update_created(tz = "Japan")
```

## Optional metadata

### Licenses

Camtrap DP expects at least one license for the data and one for the media.

```{r}
datapackage$add_license(name = "CC-BY-4.0",
                        path = "http://creativecommons.org/licenses/by/4.0/",
                        scope = "data")
datapackage$add_license(name = "CC-BY-4.0",
                        path = "http://creativecommons.org/licenses/by/4.0/",
                        scope = "media")
```

### Related identifiers

```{r}
datapackage$add_relatedIdentifiers(
  relationType          = "IsSupplementTo",
  relatedIdentifier     = "https://doi.org/xxxx",
  relatedIdentifierType = "DOI",
  resourceTypeGeneral   = "JournalArticle")
```

### Properties, sources and references

```{r}
datapackage$set_properties(
  name     = "dummy-nies",
  homepage = "https://www.nies.go.jp/biology/snapshot_japan/index.html")
datapackage$add_sources(title = "DummyData")
datapackage$add_references(reference = "DummyNIES https://doi.org/xxxxx")
```

### Custom resources

`set_custom()` attaches an extra resource (for example data used by an
abundance estimator) as metadata. It must be called after the three core tables
have been set.

```{r}
RD <- data.frame(id = seq_len(388), Time = sample(1:29, 388, replace = TRUE))
```
```{r, eval = FALSE}
datapackage$set_custom(name = "rest",
                       description = "data for the REST method",
                       data = RD)
```

# 5. Output the data package

```{r, eval = FALSE}
# return the camtrapdp object
data_camtrapdp <- datapackage$out_camtrapdp()

# or also write deployments.csv / media.csv / observations.csv + datapackage.json
datapackage$out_camtrapdp(write = TRUE, directory = path)
```

When written, the CSV files contain every schema column, booleans are written as
`true`/`false`, and unset metadata is omitted so that empty placeholders do not
cause spurious validation errors.

# 6. Validate the written package with Frictionless

## Conformance pre-checks (before calling Python)

Before running Python, you can check on the R side whether the package is even a
well-formed Frictionless data package — and whether it is Camtrap DP form. This
mirrors, in R, the structural checks Frictionless performs, so problems with a
brand-new or unusual schema surface early.

```{r, eval = FALSE}
datapackage$check_descriptor()        # package + table-schema structure (Frictionless spec)
datapackage$check_camtrap_profile()   # warn if the profile is not a Camtrap DP profile
```

A package can be a valid *Frictionless* data package without being *Camtrap DP*
form: that depends on whether its `profile` is the Camtrap DP profile (which is
the default). The authoritative check, including GeoJSON validity and the
physical file structure, is still the Frictionless run below.

## Run Frictionless

You can confirm the written package against the official schemas with the Python
[Frictionless](https://framework.frictionlessdata.io/docs/guides/validating-data.html)
validator. This requires Python with `frictionless` installed
(`pip install frictionless`).

```{r, eval = FALSE}
issues <- datapackage$validate_frictionless(directory = path, python = "python")
ctdp_is_valid(issues)   # TRUE if there are no errors
```

**Note — this rewrites `path`.** `validate_frictionless()` defaults to
`write = TRUE`, so it calls `out_camtrapdp()` and **overwrites** the
`datapackage.json` and CSVs in `directory` from the current object before
validating. To validate a package that already exists on disk **without
overwriting it**, use `write = FALSE`, or the standalone validate-only function
(no R6 object needed):

```{r, eval = FALSE}
ctdp_validate_frictionless("path/to/existing/package", python = "python")
```

`issues` is a tidy table with one row per problem, giving the `source` file, the
`field` (column or property path), the `row`, the violated `constraint`, the
offending `value`, and a `message`, so you can see exactly where any error
occurs. For cell errors `value` is the failing cell; for metadata (profile)
errors it is resolved from `datapackage.json` via the property path in the note
(e.g. `contributors[].email` → the actual email value(s)). You can also aggregate the R-side schema
checks, the relation checks, the metadata (profile) checks, the conformance
pre-checks and (optionally) the Frictionless report in one call:

```{r, eval = FALSE}
datapackage$validate(relations = TRUE, metadata = TRUE, conformance = TRUE,
                     frictionless = TRUE, directory = path, python = "python")
```

# 7. Converting an arbitrary spreadsheet directly

The helpers above assume you already named your variables. If instead you have a
raw spreadsheet with its own column names, you can map and validate it in one
step with `ctdp_build_table()`, which applies a column mapping, merges separate
date/time columns, coerces to the schema types and validates — for any version.

```{r, eval = FALSE}
version    <- "1.0.1"
dep_schema <- TableSchema$new("deployments", version = version)

# an example raw sheet with arbitrary column names + a custom column
raw <- data.frame(
  station   = c("A01", "A02"),
  lat       = c(35.1, 36.2),
  lon       = c(139.5, 140.1),
  start_day = c("2023-04-01", "2023-04-02"),
  start_clk = c("09:00:00", "10:30:00"),
  end_day   = c("2023-05-01", "2023-05-02"),
  end_clk   = c("09:00:00", "10:30:00"),
  myNote    = c("kept as a custom column", "kept too"),
  stringsAsFactors = FALSE)

# mapping: names are SOURCE columns, values are Camtrap DP FIELD names
mapping <- c(station = "deploymentID", lat = "latitude", lon = "longitude")

built <- ctdp_build_table(
  dep_schema, raw, mapping = mapping,
  datetime_merges = list(
    list(date_col = "start_day", time_col = "start_clk", target = "deploymentStart"),
    list(date_col = "end_day",   time_col = "end_clk",   target = "deploymentEnd")))

ctdp_summarize_validation(built$issues)   # any schema problems
datapackage$set_deployments(built$data)   # feed the result into the package
```

Custom columns such as `myNote` are kept; when the package is written, the
custom column is declared in an inline extended schema in `datapackage.json` so
that Frictionless accepts it.

# 8. Other schema flavors (e.g. bioacoustics)

Because every table is driven by the schema you point it at, the package is not
limited to the camera-trap schemas hosted by TDWG. To target a different flavor —
for instance the
[bioacoustics extension](https://github.com/camera-traps/bioacoustics) of
Camtrap DP — give the table and profile URLs explicitly. These schemas live in a
different repository and use their own field set (e.g. `deviceID` instead of
`cameraID`, plus `samplingFrequency`, `frequencyLow`/`frequencyHigh`, ...) and
per-table datetime formats (the `media` / `observations` event timestamps use
fractional seconds `%Y-%m-%dT%H:%M:%S.%f%z`, while the `deployments` times do
not); the schema-driven validation adapts to all of this automatically. If your
raw `media` / `observations` timestamps lack the fractional part, `.000` is
added automatically so the value matches the schema's `%f` format.

Point the package at the flavor once with `set_properties()`, then add tables as
usual — the `set_*()` methods use the configured `schema_urls`, so you do not
need to pass `schema =` to each call:

```{r, eval = FALSE}
ba <- "https://raw.githubusercontent.com/camera-traps/bioacoustics/main/camtrap-dp/1.0.2/%s"

dp <- R6_CamtrapDP$new(version = "1.0.2")
dp$set_properties(
  version     = "1.0.2",
  profile     = sprintf(ba, "camtrap-dp-profile-acoustic.json"),
  schema_urls = list(
    deployments  = sprintf(ba, "deployments-table-schema-acoustic.json"),
    media        = sprintf(ba, "media-table-schema-acoustic.json"),
    observations = sprintf(ba, "observations-table-schema-acoustic.json")))

# audio timestamps carry fractional seconds to match the acoustic schema format
dp$set_media(data.frame(
  mediaID = "m1", deploymentID = "D1",
  timestamp = "2023-04-01T09:05:00.000+0900",
  filePath = "audio/m1.wav", filePublic = TRUE, fileMediatype = "audio/wav",
  samplingFrequency = 48000L, channels = 1L,
  stringsAsFactors = FALSE))
```

## Mapping camera-trap columns to the acoustic flavor

You only need a mapping for columns whose **name differs** from the acoustic
field. Columns that already use the acoustic field name (`deploymentID`,
`latitude`, `deploymentStart`, ...) are matched automatically — no mapping
needed. For deployments, the camera-trap `camera*` fields are renamed to
`device*`; the camera-only fields have no acoustic equivalent and should be
dropped; and a few acoustic-only fields can be set if you have the data.

```{r, eval = FALSE}
library(dplyr)

# camera-trap deployments -> acoustic deployments (only the renamed columns)
mapping <- c(
  cameraID      = "deviceID",
  cameraModel   = "deviceModel",
  cameraDelay   = "deviceDelay",
  cameraHeight  = "deviceHeight",
  cameraDepth   = "deviceDepth",
  cameraTilt    = "deviceTilt",
  cameraHeading = "deviceHeading")

dep_acoustic <- camtrap_deployments %>%
  select(-any_of(c("featureType", "timestampIssues")))   # camera-only: no acoustic field

dp$set_deployments(dep_acoustic, mapping = mapping)
```

Field correspondence — **deployments**:

| Camera-trap field | Acoustic field | Action |
|---|---|---|
| `deploymentID`, `locationID`, `locationName`, `latitude`, `longitude`, `coordinateUncertainty`, `deploymentStart`, `deploymentEnd`, `setupBy`, `detectionDistance`, `baitUse`, `habitat`, `deploymentGroups`, `deploymentTags`, `deploymentComments` | *same name* | **no mapping** |
| `cameraID` / `cameraModel` / `cameraDelay` / `cameraHeight` / `cameraDepth` / `cameraTilt` / `cameraHeading` | `deviceID` / `deviceModel` / `deviceDelay` / `deviceHeight` / `deviceDepth` / `deviceTilt` / `deviceHeading` | **map** |
| `featureType`, `timestampIssues` | — | **drop** |
| — | `elevation`, `devicePlatform`, `recordingSchedule`, `locationType` | acoustic-only (set if available) |

For **observations** the only renamed field is `cameraSetupType` → `deviceSetupType`
(acoustic also adds `frequencyLow` / `frequencyHigh` / `classificationConfirmation`).
For **media** there are no renames, only extra fields (`duration`, `bitDepth`,
`samplingFrequency`, `gain`, `channels`).

Inspect a flavor the same way as any other schema. Note that
`TableSchema$new("deployments", version = "1.0.2")` **without** `url_template`
loads the *camera-trap* deployments schema; pass the acoustic URL to inspect the
acoustic requirements. `requirements()` returns a tidy table of every field's
type, format and constraints.

```{r, eval = FALSE}
acoustic_dep <- TableSchema$new(
  "deployments", version = "1.0.2",
  url_template = sprintf(ba, "deployments-table-schema-acoustic.json"))
acoustic_dep$field_names()
acoustic_dep$required_field_names()
acoustic_dep$requirements()        # field / type / format / required / enum / min / max / pattern
acoustic_dep$external_references()
```

> Note that `create_deployments()`, `create_media()` and `create_observations()`
> are tailored to the camera-trap schema. For a different flavor (or for new
> columns in a future version), build the tables with the schema-driven path
> (`ctdp_build_table()` or the `set_*()` methods with a custom `schema =`)
> rather than the `create_*()` helpers.
