---
title: "R2camtrapdp: acoustic (audio) data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R2camtrapdp acoustic (audio) data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(R2camtrapdp)
```

# Overview

Acoustic (audio) recordings use the **bioacoustics flavor** of Camtrap DP. Audio
data is **media-based** (`observationLevel = "media"`): each observation refers
to a media file (`mediaID`). The differences from the camera-trap schema are read
automatically from the bioacoustics schemas; the main ones are:

* **deployments** use **`device*`** fields (`deviceID`, `deviceModel`, ...) plus
  `elevation`, `devicePlatform`, `recordingSchedule`, `locationType`.
* **media** add `duration`, `samplingFrequency`, `bitDepth`, `gain`, `channels`;
  `fileMediatype` is an audio type such as `audio/wav`.
* **observations** add `frequencyLow`, `frequencyHigh`; `cameraSetupType` becomes
  `deviceSetupType`.
* **datetime formats differ per table**: `deployments` use `%Y-%m-%dT%H:%M:%S%z`
  (no fractional seconds); `media` / `observations` timestamps use fractional
  seconds `%Y-%m-%dT%H:%M:%S.%f%z`.
* `project$captureMethod` for audio is one of `activityDetection`, `continuous`,
  `recordingSchedule` (not `timeLapse`).

> **Tip — pass datetimes as `POSIXct`.** The package then formats each table's
> datetime correctly (offset `+0900`, and `.000` fractional seconds where the
> schema requires them). A raw string like `"2026/6/12 12:00:00"` is written
> as-is and fails validation.

This example assumes you have only two field notebooks — a **deployment**
notebook and an **observation** notebook — and that the observation notebook
carries the audio **file names**, from which `media` is derived.

# Data

```{r}
data("Adep")   # deployment field-notebook (one row per device deployment)
data("Aobs")   # observation field-notebook (one row per observation; has `filename`)
str(Adep, vec.len = 2)
str(Aobs, vec.len = 2)
```

`Adep` has `deploymentID`, coordinates, `startDate`/`startTime`,
`endDate`/`endTime`, `deviceID`, `deviceModel`, the recording settings
(`samplingFrequency`, `bitDepth`, `channels`) and `setupBy`. `Aobs` has
`deploymentID`, `filename`, `date`/`time`, `duration`, the taxonomy
(`class`/`genus`/`species`), `individualCount`, `frequencyLow`/`frequencyHigh`
and `eventStart`/`eventEnd`. In this example data the coordinates fall in the
Izu Peninsula (Japan), `individualCount` is `NA` (not counted from audio), and
`frequencyLow`/`frequencyHigh` use approximate values from the literature.

# 1. Point the package at the bioacoustics flavor

```{r, eval = FALSE}
ba <- "https://raw.githubusercontent.com/camera-traps/bioacoustics/main/camtrap-dp/1.0.2/%s"

dp <- R6_CamtrapDP$new(version = "1.0.2",
  title = "Acoustic survey example", description = "AudioMoth recordings",
  id = "https://example.org/dataset/acoustic-1")

dp$set_properties(
  version     = "1.0.2",
  profile     = sprintf(ba, "camtrap-dp-profile-acoustic.json"),
  schema_urls = list(
    deployments  = sprintf(ba, "deployments-table-schema-acoustic.json"),
    media        = sprintf(ba, "media-table-schema-acoustic.json"),
    observations = sprintf(ba, "observations-table-schema-acoustic.json")))
```

# 2. Deployments (from the deployment notebook)

Build the deployments table from `Adep`. Combine the date and time columns into
`POSIXct` so the package writes the correct datetime format.

```{r}
deployments <- data.frame(
  deploymentID    = Adep$deploymentID,
  latitude        = Adep$latitude,
  longitude       = Adep$longitude,
  locationID      = Adep$locationID,
  deploymentStart = as.POSIXct(paste(Adep$startDate, Adep$startTime), tz = "Japan"),
  deploymentEnd   = as.POSIXct(paste(Adep$endDate,   Adep$endTime),   tz = "Japan"),
  deviceID        = Adep$deviceID,
  deviceModel     = Adep$deviceModel,
  setupBy         = Adep$setupBy,
  stringsAsFactors = FALSE)
```
```{r, eval = FALSE}
dp$set_deployments(deployments)
```

# 3. Media (derived from the observation notebook's file names)

There is no separate media notebook: build `media` from the unique `filename`s
in `Aobs` (one row per audio file), and bring the recording settings over from
`Adep`.

```{r}
files <- Aobs[!duplicated(Aobs$filename), ]   # one row per audio file

media <- data.frame(
  mediaID       = files$filename,             # the file name is the media identifier
  deploymentID  = files$deploymentID,
  timestamp     = as.POSIXct(paste(files$date, files$time), tz = "Japan"),
  filePath      = file.path("audio", files$filename),
  filePublic    = TRUE,
  fileMediatype = paste0("audio/", tolower(tools::file_ext(files$filename))),  # "audio/wav"
  duration      = files$duration,
  stringsAsFactors = FALSE)

# add device recording settings (samplingFrequency / bitDepth / channels) from Adep
media <- merge(media, Adep[, c("deploymentID", "samplingFrequency", "bitDepth", "channels")],
               by = "deploymentID", all.x = TRUE)
head(media)
```
```{r, eval = FALSE}
dp$set_media(media)
```

`timestamp` is written **with** fractional seconds (e.g.
`2026-06-12T06:00:00.000+0900`) to match the acoustic `media` format — handled
automatically because it is a `POSIXct`.

# 4. Observations (from the observation notebook)

```{r}
observations <- data.frame(
  observationID    = paste(Aobs$deploymentID, Aobs$eventID, Aobs$obsID, sep = "_"),
  deploymentID     = Aobs$deploymentID,
  mediaID          = Aobs$filename,           # link to media (mediaID = filename)
  eventStart       = as.POSIXct(Aobs$eventStart, tz = "Japan"),
  eventEnd         = as.POSIXct(Aobs$eventEnd,   tz = "Japan"),
  observationLevel = "media",
  observationType  = ifelse(Aobs$object == "none", "blank",
                     ifelse(Aobs$object == "hito", "human", "animal")),
  scientificName   = ifelse(is.na(Aobs$genus), Aobs$class, paste(Aobs$genus, Aobs$species)),
  count            = Aobs$individualCount,    # NA here (not counted from audio)
  frequencyLow     = Aobs$frequencyLow,
  frequencyHigh    = Aobs$frequencyHigh,
  stringsAsFactors = FALSE)
```
```{r, eval = FALSE}
dp$set_observations(observations)
```

# 5. Metadata, relations, write, validate

```{r, eval = FALSE}
dp$add_contributors(data.frame(title = "Jane Doe", role = "contact",
                               organization = "NIES", stringsAsFactors = FALSE))
dp$add_license(name = "CC0-1.0",   scope = "data")
dp$add_license(name = "CC-BY-4.0", scope = "media")
dp$set_project(title = "Acoustic survey", samplingDesign = "systematicRandom",
               captureMethod = "recordingSchedule", individualAnimals = FALSE,
               observationLevel = "media")
dp$set_st()
# dp$set_taxon()   # taxonID from GBIF/ITIS/NCBI; requires the taxadb package + internet

dp$check_relations()   # PK/FK; warns and points at datapackage$data$... if a key is missing

path <- file.path(tempdir(), "acoustic-package")
dp$out_camtrapdp(write = TRUE, directory = path)

issues <- dp$validate_frictionless(directory = path, python = "python")  # pip install frictionless
ctdp_is_valid(issues)
```

To validate a package that already exists on disk **without overwriting it**:

```{r, eval = FALSE}
ctdp_validate_frictionless("path/to/existing/acoustic-package", python = "python")
```

# 6. Inspecting the acoustic requirements

```{r, eval = FALSE}
ba <- "https://raw.githubusercontent.com/camera-traps/bioacoustics/main/camtrap-dp/1.0.2/%s"
acoustic_media <- TableSchema$new(
  "media", version = "1.0.2",
  url_template = sprintf(ba, "media-table-schema-acoustic.json"))

acoustic_media$field_names()
acoustic_media$requirements()                 # type / format / required / enum per field
acoustic_media$field("timestamp")$format      # "%Y-%m-%dT%H:%M:%S.%f%z"
```
